---
title: "PCA in R"
output: html_notebook
---

<details>
  <summary>Function Docs Refresher</summary>

- `matrix()`: creates a matrix from the given set of values
  - we may provide no data (NA), and just specify `nrow = 5`, `ncol = 5`
- `colnames(x)`: where `x` is a matrix-like R object, with at least two dimensions for colnames
- `paste0()`: concatenate vectors after converting to character
- `sample(x, size, replace, prob)`: takes a sample of the specified size from the elements of `x` using either with or without replacement
  - `x`: a vector of values from which the sample will be drawn
  - `n`: a positive number, the number of items to choose from
  - `size`: number of items to draw from `x`
  - `replace`: should sampling be with replacement?
  - `prob`: a vector of probability weights for obtaining the elements of the vector being sampled
  - Example:
    - `x = 10:1000`: This creates a sequence of numbers starting from 10 and ending at 1000. So, x is a vector containing numbers [10, 11, 12, ..., 1000]
    - `size = 1`: This means we want to draw one number from the sequence 10:1000
    - Given these parameters, sample() will randomly select one value from the sequence of numbers between 10 and 1000 (inclusive).
- `rpois(n, lambda)`:

</details>


### What we cover:
1. How to use the `prcomp()` function to do PCA
2. How to draw a PCA plot using *base graphics* and *ggplot2*
3. How to determine how much variation each principal component accounts for
4. How to examine the loading scores to determine what variables have the largest effect on the graph

### Generate a dataset
First, let's generate a fake dataset that we can use in the demonstration

- `wt` - wild type - samples are the normal, every day samples
- `ko` - knock-out - samples are samples that are missing a gene because we knocked it out

```{r}
data.matrix <- matrix(nrow = 100, ncol = 10)

# name the samples
colnames(data.matrix) <- c( # combines the two vectors, assign them as clumn names
  paste0("wt", 1:5), # creates wt1...wt5
  paste0("ko", 1:5) # creates ko1...ko5
)

# name the genes
rownames(data.matrix) <- paste0("gene", 1:100)

# give the fake genes fake read counts
for (i in 1:100) {
  wt.values <- rpois(5, lambda = sample(x = 10:1000, size = 1))
  ko.values <- rpois(5, lambda = sample(x = 10:1000, size = 1))

  data.matrix[i,] <- c(wt.values, ko.values)
}

head(data.matrix)
```

### Doing PCA

- The goal is to draw a graph to show how the samples are related (or not related) to each other
- By default `prcomp()` expects the samples to be rows and the genes to be columns
- Since the samples in our data matrix are columns, and the genes (variables) are rows we have to transpose the matrix using the `t()` function
- `prcomp()` returns three things:
  1. `x`
    Contains the principal components (PCs) for drawing a graph. Here we are using the first two columns in `x` to draw a 2-D plot that uses the first two PCs. Even though overall there are 10 PCs (since there are 10 samples).
    The first PC accounts for the most variation in the original data, the second PC for the second most and so on. We use frst two PCs to plot a 2D PCA graph. However, sometimes we use PC2 and PC3.
  2. `sdev`
     Stands for "standard deviation". We use it to calculate how much variation in the original data each principal component accounts for.
  3. `rotation`

- PC1 is on the x-axis, because it is the first column in `x`
- PC2 is on the y-axis

To get a sense of how meaningful these clusters are, let's see how much variation in the original data PC1 accounts for. For that we use the square of `sdev`.

Looking at the plot with calculated variations we can see that PC1 accounts for almost all the data variation in the data. This means that there is a big difference between those two clusters.

Breakdown of `ggplot()`:

1. Pass in the `pca.data` dataframe and tell **ggplot** which columns contain the X and Y coordinates and which column has the sample labels
2. `geom_text()` - plot the labels, rather than dots or some other shape
3. `xlab()` and `ylab()` to add X and Y axis labels.

Lastly, let's look at how to use **loading scores** to determine which genes have the largest effect on where samples are plotted in the PCA plot.

- There are loading scores for each PC
- We are going to look at the loading scores of PC1, because it accounts for 92% of the variation in the data
- Genes that push samples to the left of the graph will have large negative values and genes that push samples to the right will have large positive values
- Since we're interested in both sets of genes, we'll use the `abs()` function to sort based on the number's magnitude rather than from high to low

```{r}
pca <- prcomp(t(data.matrix), scale = TRUE)
plot(pca$x[,1], pca$x[,2])

# Calculate variation
pca.var <- pca$sdev^2

# Calculate the percentages
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

# Plot the percentages
barplot(pca.var.per, main = "Scree plot", xlab = "Principal Component", ylab = "Percentage Variation")

library(ggplot2)

pca.data <- data.frame(Sample=rownames(pca$x),
                       X=pca$x[,1],
                       Y=pca$x[,2])
pca.data

ggplot(data = pca.data, aes(x=X, y=Y, label=Sample)) +
        geom_text() +
        xlab(paste0("PC1 - ", pca.var.per[1], "%")) +
        ylab(paste0("PC2 - ", pca.var.per[2], "%")) +
        theme_bw() +
        ggtitle("PCA Graph")

# Loadinf scores
loading_scores <- pca$rotation[,1]
gene_scores <- abs(loading_scores)
gene_scores_ranked <- sort(gene_scores, decreasing = TRUE)
top_10_genes <- names(gene_scores_ranked[1:10])

top_10_genes

pca$rotation[top_10_genes, 1] # show the scores (now with +/- sign)
```

